
\chapter{Related Work}
\label{chapt:related-work}

The most closely related work to CubedOS is NASA's Core Flight System \cite{cFE}. Like CubedOS,
CFS endeavors to be a general purpose framework for building flight software. Also like CubedOS,
CFS is associated with a collection of modules that support common functionality needed by many
missions. In addition, CFS makes use of a message passing discipline using a publish/subscribe
model of message handling. CubedOS can provide support for publish/subscribe message handling by
way of a library module.

The main difference between the systems, aside from maturity level (CFS is a long established
project with a history of actual use), is that CubedOS is written in \SPARK\ and verified free
of runtime error. In contrast, CFS is written in C with no particular static verification goals.

The CFS architecture is layered, whereas CubedOS modules operate as peers of each other. The CFS
architecture makes use of a separate Operating System Abstraction Layer (OSAL) that presents a
common interface across all the platforms supported by CFS. In contrast, CubedOS relies on the
Ada runtime system for this purpose, and is thus Ada specific, but remains operating system
independent. CubedOS also uses a module library, the CubedOS LLAL, to provide hardware and OS
independence in areas not covered by the Ada runtime system and standard library, but the
interface to these modules is not yet standardized.

Kubos \cite{kubos} is a project with roughly similar goals as CFS and CubedOS. It is not as
mature as CFS. Like CFS, Kubos is written in C without static verification goals in the sense
that we mean here.

Some CubeSat flight software is written directly on top of conventional embedded operating
systems such as Linux, VxWorks \cite{vxworks}, or RTEMS \cite{rtems}. These systems allow
application software to potentially be written with a variety of tools and methods, although C
is most often used in practice. They provide flexibility by imposing few restrictions, but they
also don't, by themselves, provide support for common flight software needs. Also they are
themselves not statically verified as CubedOS is, although the Wind River VxWorks Cert platform
\cite{vxworks-cert} does provide a means by which VxWorks can be used in safety critical
avionics applications conforming to the DO-178B standard.
