%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FILE    : main.tex
% AUTHOR  : (C) Copyright 2018 by Sean Klink & Carl Brandon & Peter Chapin
% SUBJECT : Paper on CubedOS
%
% TODO: Write me! Add more specific items here as they become apparent.
%
% Send comments or bug reports to:
%
%    Carl Brandon
%    Vermont Technical College
%    Randolph Center, VT 05061
%    CBrandon@vtc.vsc.edu
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%+++++++++++++++++++++++++++++++++
% Preamble and global declarations
%+++++++++++++++++++++++++++++++++
\documentclass{llncs}
\usepackage{listings}
\usepackage{gensymb}
\usepackage{graphicx}
%\usepackage{unicode-math}
\usepackage{url}
\usepackage{hyperref}

\input{macros}

%++++++++++++++++++++
% The document itself
%++++++++++++++++++++
\begin{document}

\title{Message Encoding for CubedOS Using the XDR2OS3 Code Generator}
\author{Sean Klink\inst{1} \and Peter Chapin\inst{3} \and Carl Brandon\inst{2} }

\institute{Vermont Technical College, Randolph Center VT 05061, USA,\\
  \email{sean.klink@vtc.edu} %,\\
      % WWW home page: \texttt{http://web.vtc.edu/users/xyzzy}
  \and
  \email{carl.brandon@vtc.edu} %,\\
      % WWW home page: \texttt{http://web.vtc.edu/users/xyzzy}
  \and
  Vermont Technical College, Williston VT 05495, USA,\\
  \email{pchapin@vtc.vsc.edu} %,\\
      % WWW home page: \texttt{http://web.vtc.edu/users/pcc09070}
}

\maketitle

\begin{abstract}
  % Summarize the contents of the paper using no more than 400 words.

  We are developing the flight software the Lunar IceCube mission, scheduled to be launched in
  2018. Lunar IceCube will be a 6-U CubeSat orbiting the moon in search of water, ice, and other
  volatiles near the lunar surface. Lunar IceCube will be the first spacecraft to use CubedOS, a
  general purpose, open source application framework for CubeSat flight software written by us.
  CubedOS is written in \SPARK\ 2014 and proved free of certain classes of runtime error. It
  uses a message passing architecture where each concurrently executing module processes
  messages from an associated mail box and sends messages to other mail boxes. CubedOS
  encourages a client/server software architecture where server modules process request messages
  and return reply messages to clients. Message data is encoded using the external data
  representation (XDR) standard. Each module thus provides encoding and decoding subprograms so
  the clients of a module are not required to manage message encodings themselves. These
  subprograms are tedious for module authors to produce manually. In this paper we present
  XDR2OS3, a tool that enables CubedOS module authors to specify message structures using an
  interface language that follows the XDR standard, extended to account for the special
  properties of \SPARK\ (such as constrained subtypes). XDR2OS3 uses the message structures and
  generates a \SPARK\ package containing suitable encoding and decoding subprograms. The output
  of XDR2OS3 can be analyzed by the \SPARK\ tools and proved free of runtime error. While the
  development of XDR2OS3 is a work in progress, our preliminary experience with the tool is
  encouraging. It allows module authors to focus on message types and how they will be used and
  supported rather than on the low level details of message formatting. We also discuss the
  potential for XDR2OS3 to be used in other \SPARK\ applications with an interest in XDR encoded
  data.
  
  \keywords{SPARK, student project, CubeSat}
\end{abstract}

\input{introduction}
\input{conclusion}

\bibliographystyle{splncs03}
\bibliography{../../LaTeX/references-General,../../LaTeX/references-SPARK,../../LaTeX/references-CubeSat}

\end{document}
