%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FILE    : main.tex
% AUTHOR  : (C) Copyright 2015 by Vermont Technical College
% SUBJECT : Potential paper on CubedOS
%
% TODO:
%
% Send comments or bug reports to:
%
%    Carl Brandon
%    Vermont Technical College
%    Randolph Center, VT 05061
%    carl.brandon@vtc.edu
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%+++++++++++++++++++++++++++++++++
% Preamble and global declarations
%+++++++++++++++++++++++++++++++++
\documentclass[twocolumn]{article}
\usepackage{authblk}
\usepackage[pdftex]{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{url}

\lstdefinelanguage{Ada2012}    %Define listing parameters for Ada 2012
   [2005]{Ada}                 %... base it on Ada 2005
   {morekeywords={some},
    sensitive=false,
    breaklines=false,
    showstringspaces=false,
    basicstyle=\small\sffamily,
    columns=flexible,
    }
\lstset{language=Ada2012, showlines=true}

\newcommand{\SPARK}{\textsc{Spark}}   % for ease of small capping SPARK

% Stock article class title page formatting...
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{HIGH INTEGRITY SOFTWARE FOR CUBESATS AND OTHER SPACE MISSIONS}
\author[*]{Carl Brandon}
\author[**]{Peter Chapin}
\affil[*]{Vermont Technical College, Science Department}
\affil[**]{Vermont Technical College, Computer Information Systems Department}
\renewcommand\Authands{ and }
\date{}

%++++++++++++++++++++
% The document itself
%++++++++++++++++++++
\begin{document}

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
\begin{abstract}
  % Summarize the contents of the paper using at least 70 and at most 400 words.

  We currently have an operating single CubeSat launched as part of NASA's ELaNa IV program on
  November 19, 2013, the first satellite of any kind launched by a college or university in New
  England. Many CubeSat failures have been attributed to software failures. Of the twelve
  university CubeSats that were launched with ours, we are the only one that is fully
  functional. One damaged their batteries the first day due to a software error and worked
  partially for a short time only in sunlight, one had partial contact for a week, one lasted
  four months, and eight were never heard from. These other CubeSats primarily used the C
  language. We are using the most reliable software technology ever sent into space. We used the
  \SPARK\ 2005 Toolset and Ada language in the construction of our software. Ada is used in
  almost all European Space Agency and many NASA rockets and spacecraft, and in most European
  rail systems and nuclear power plants. \SPARK\ is used in commercial aviation (Rolls-Royce
  Trent jet engines, ARINC ACAMS system, Lockheed Martin C130J), military aviation (EuroFighter
  Typhoon, Harrier GR9, AerMacchi M346), air-traffic management (UK NATS iFACTS system), rail
  (numerous signaling applications), and medical (LifeFlow ventricular assist device)
  applications.

  We are using \SPARK/Ada, with its reduction of errors by a factor of about 100 compared with
  C. \SPARK\ is a formally defined programming language and a set of verification tools
  specifically designed to support the development of high integrity software, and can formally
  verify information flow, freedom from runtime errors, functional correctness, and security and
  safety policies.

  Ours is the first spacecraft to use \SPARK. We are currently upgrading our CubeSat software to
  \SPARK\ 2014, and will then work on improving some of the algorithms in that software. We
  would then have a very reliable software platform, CubedOS, that other projects could use a a
  base for their CubeSat or other spacecraft projects. We have also been converting NASA
  Goddard's GPS Enhanced Onboard Navigation System (GEONS) to \SPARK, and have discovered and
  fixed errors by this process. Our next CubeSat, Lunar IceCube with Morehead State University
  (PI), Goddard Space Flight Center (BIRCHES \& Lunar transfer trajectory)and the Jet Propulsion
  Lab (Iris 2) is self propelled with a Busek iodine ion drive which will go to the Moon on the
  Space Launch System EM-1 flight in 2018. This software will be much more complex, dealing with
  power management, the ADACS, the infrared spectrometer (BIRCHES), the data and navigation
  radio (Iris 2), the electrical power system, and aim the photo voltaic panels and ion
  thruster. The software will carry out the navigation plan and deal with ground based commands
  and upgrades. \SPARK’s reliability will be necessary for this.

  %\keywords{\SPARK, student project, CubeSat}
\end{abstract}
  \end{@twocolumnfalse}
]

%\begin{enumerate}
%\item Vermont Lunar CubeSat Description
%  \begin{enumerate}
%  \item Hardware
%  \item Software
%    \begin{enumerate}
%      \item Software Metrics
%    \end{enumerate}
%  \item Mission
%  \item ELaNa IV Results
%  \end{enumerate}
%
%\item \SPARK\ 2014 Description and Characteristics
%\begin{enumerate}
%  \item Toolset
%  \item Language
%  \item Where \SPARK\ would have helped
%    \begin{enumerate}
%    \item Ariane V
%    \item Boeing 787
%    \end{enumerate}
%  \end{enumerate}
%
%\item New Work
%\begin{enumerate}
%  \item CubedOS
%  \item GEONS translation
%  \item Lunar IceCube
%    \begin{enumerate}
%    \item Collaborators
%    \item Spacecraft description
%      \begin{enumerate}
%      \item Iodine ion drive
%      \item Computer
%      \item BIRCHES
%      \item Iris 2
%      \item ADAC
%      \item PV panels
%      \end{enumerate}
%    \item Mission
%      \begin{enumerate}
%      \item Transfer to lunar orbit
%      \item Collecting data
%      \item Data download
%      \item End - Moon or Mars?
%      \end{enumerate}
%    \end{enumerate}
%  \end{enumerate}
%\end{enumerate}

\input{introduction}
\input{vtlunar}
\input{spark2014}
\input{new}
\input{conclusion}

\bibliographystyle{plain}
\bibliography{../../../LaTeX/references-CubeSat,../../../LaTeX/references-SPARK}

\end{document}
